Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis

### 2018-05-28 rigid_bodies.tex
added supplementary material

% Comment below before submitting
\let\titlefont\undefined
\usepackage[fontsize=11pt]{scrextend}
\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
%\flushbottom
% Comment above before submitting

\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

\usepackage{array}
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{C}{>{$}c<{$}}
\newcolumntype{R}{>{$}r<{$}}

@@

\subsection{NVE Dynamics: Notation and Symplectic Integration}

In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.

@@

%The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the body-fixed frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).

@@

\subsection{Measure-Preserving Integration for the NVT Dynamics}

As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
Ezra \cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion.
In short, a splitting scheme will preserve a given measure if each individual propagator does it as well.
By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
Of course, the NVE part can be integrated exactly as described before.
The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
In the following, the NVT integrator introduced in the present section will be denoted by MKKT as its development was inspired by Ref.~\cite{Martyna_1996}.

@@

\subsection{Backward Error Analysis}

In order to explain our proposal, we turn the attention again to the NVE case.
A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian.
In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors).
It is also known that $\Ham{S}$ depends on the time step size $\timestep$.
As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right),
\end{equation}
where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the resultant torque vector.
Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
In turn, the modified kinetic energy can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$.
Therefore, a Trotter-Suzuki trajectory, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\modified K}{\vt p}, \;
\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q},
\end{equation*}
where $\modified{\Ham{}} = \modified K + \modified U$ is a modified Hamiltonian.
The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
\end{equation}

Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained Trotter-Suzuki trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

\subsection{Adjustment of the NVT Dynamics}

Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly as described before.
The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
\begin{equation}
G_1 = 2 \modified K - L k_\text{\tiny B} T.
\end{equation}

One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.

@@

@@

\section{Theoretical Background}
\label{sec:theory}

\subsection{Hamiltonian Dynamics}
\label{eq:hamiltonian dynamics}

When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of the center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t)$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}.
In this way, the connection between the body-fixed and space-fixed frames of reference can be recovered at any time by the rotation matrix $\mt A(\vt q)$, which admits the factorization mentioned in Sec.~\ref{sec:methodology}.
This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$.
Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
\begin{equation*}
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\end{equation*}

The intermediate frame is exactly the subspace of $\mathbb{R}^4$ that must contain the rate-of-change vector $\dot{\vt q}$ if the norm of $\vt q$ is to remain constant, since $\tr{\vt q} {\vt q} = 1$ implies that $\tr{\vt q} \dot{\vt q} = 0$.
In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.

The Liouville operator that stems from the Hamiltonian in Eq.~\eqref{eq:Hamiltonian} is\cite{Silveira_2017}
\begin{equation}
\label{eq:Lioville operator}
\begin{split}
\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \tr{\vt F} \diff{}{\vt p} \\
&+ \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi}.
\end{split}
\end{equation}

Because the time evolution of any phase-space vector $\vt x$ is governed by $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ \cite{Tuckerman_2010}, the equations of motion that describe the rigid-body motion are those in Eq.~\eqref{eq:ODE system for NVE}.
The solution for a system like $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ with initial condition $\vt x(0) = \vt x_0$ can be formally expressed as $\tr{\vt x}(t) = e^{t \Liu{\tiny NVE}} \tr{\vt x}_0$.
Although the exact effect of the exponential operator %$e^{t \Liu{\tiny NVE}}$
cannot be determined for an arbitrary potential $U(\vt r, \vt q)$, a symplectic numerical solution can be achieved by a Verlet-type integration.
For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
\begin{equation}
\label{eq:symmetric BCH}
\begin{split}
&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \\
&= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
\end{split}
\end{equation}
where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
With $\Liu{\tiny NVE} = \Liu{A} + \Liu{B}$ and a small value for $\timestep$, we can approximate the BCH equation by the Trotter-Suzuki splitting formula, Eq.~\eqref{eq:splitting NVE}.
In the rigid-body case, by making
\begin{subequations}
	\label{eq:Liouvile operators K and U}
	\begin{equation*}
	\label{eq:Lioville operator K}
	\Liu{A} = \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) \diff{}{\vt \pi}
	\end{equation*}
	and
	\begin{equation*}
	\label{eq:Lioville operator U}
	\Liu{B} = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \diff{}{\vt \pi},
	\end{equation*}
\end{subequations}
it becomes possible to determine exactly the individual actions of operators $e^{\timestep \Liu{A}}$ and $e^{\frac{\timestep}{2} \Liu{B}}$ because they yield two separate, analytically solvable systems of equations.
The one that originates from $e^{\timestep \Liu{A}}$ is made up of equations $\dot{\vt r} = {\mt M}^{-1} {\vt p}$, $\dot{\vt p} = {\vt 0}$, $\dot{\vt q} = \frac{1}{2} \mt B(\vt q) \vt \omega$, and $\dot{\vt \pi} = \frac{1}{2} \mt B(\vt \pi) \vt \omega$, while the one that stems from $e^{\frac{\timestep}{2} \Liu{B}}$ is composed of $\dot{\vt r} = {\vt 0}$, $\dot{\vt p} = {\vt F}$, $\dot{\vt q} = {\vt 0}$, and $\dot{\vt \pi} = 2 \mt C(\vt q) \vt \tau$.
Their solutions provides us with the propagator actions described in Sec.~\ref{sec:methodology}.

\subsection{Nos\'e-Hoover Chain Applied to Rigid Bodies}
\label{sec:canonical}

>>>>

% Comment below before submitting
\let\titlefont\undefined
\usepackage[fontsize=11pt]{scrextend}
\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
%\flushbottom
% Comment above before submitting
\usepackage{booktabs}
\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}
\usepackage{tikz}
\usetikzlibrary{decorations.text}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

\usepackage{array}
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{C}{>{$}c<{$}}
\newcolumntype{R}{>{$}r<{$}}
\usepackage{booktabs}    
\usepackage[exponent-product=\times]{siunitx}

@@

\subsection{NVE Dynamics: Notation and Symplectic Integration}
\label{sec:nve}
In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.

@@

The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).

@@

\subsection{Measure-Preserving Integration for the NVT Dynamics}
\label{sec:mttk}
As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. For that, we employed the systematic method introduced by Ezra \cite{Ezra_2006} for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion, which is 
In short, a splitting scheme will preserve a given measure if each individual propagator does it as well. 
It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure is given by
\begin{equation}
\label{eq:nhc_measure}
w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
\end{equation}

will preserve a given measure $e^{-w(\vt x)}d\vt x$

By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
Of course, the NVE part can be integrated exactly as described before.
The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
In the following, the NVT integrator introduced in the present section will be denoted as MTTK as its development was inspired by Ref.~\cite{Martyna_1996}.

@@

\subsection{Backward Error Analysis}
\label{sec:modified_h}

In order to explain our proposal, we turn the attention again to the NVE case.
A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}.
In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
It is also known that $\Ham{S}$ explicity depends on the time step size $\timestep$.

The first new result we present in this paper is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.

As shown in Appendix~\ref{sec:rigid body shadow hamiltonian}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right),
\end{equation}
where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the resultant torque vector.
Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
In turn, the modified kinetic energy can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
\begin{equation*}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right]
\end{equation*}
and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\modified K}{\vt p}, \;
\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}.
\end{equation*}

The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
\end{equation}

Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

\subsection{Adjustment of the NVT Dynamics}
\label{sec:smttk}
Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \modified K - L k_\text{\tiny B} T.
\end{equation}

One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](3,.2) -- (3,1.5);
%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-h}$};
\node at (2.2,-0.2)[below]{$r_{t}$};
\node at (4.4,-0.2)[below]{$r_{t+h}$};
\node at (6.6,-0.2)[below]{$r_{t+2h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}
%\begin{figure}
%\begin{tikzpicture}
%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](2,.2) -- (2,1.5);
%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
%\node(0) at (0,0)[shape=circle,draw, minimum width=1cm, minimum height=1cm]        {$r_{t-h}$};
%\node(1) at (2.5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_t$};
%\node(2) at (5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_{t+h}$};
%\node(3) at (7.5,0)[shape=circle,draw,minimum width=1cm, minimum height=1cm] {$r_{t+ 2h}$};
%\node(4) at (3.75,1)[] {$e^{\timestep \Liu{\tiny NVE}}$};
%\path [dashed,->]
%  (1)      edge                 node [above]  {virtual}     (0)
%  (2)      edge                 node [above] {virtual}     (3);
%\path [->,thick] 
%  (1)      edge                 node [above]  {}            (2);
%\end{tikzpicture}
%\caption{Schematic representation}
%\label{fig:virtual}
%\end{figure}

A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.

@@

$\widetilde{H}$, but an ``extended energy'' given by \cite{Martyna_1992}
\begin{equation}
\label{eq:this extended energy}
\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}

@@

\section{Theoretical Background}
